Customizing Software by Direct Manipulation of Tabular Data

### 2020-05-21 paper.tex
related work

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Customizing Software by Direct Manipulation of Structured Data}

@@

@@

\hypertarget{customization-by-direct-manipulation}{%
\subsection{Customization by direct
manipulation}\label{customization-by-direct-manipulation}}

@@

It is important to note that there are many specific customizations that
can be achieved in systems like Chickenfoot \citep{bolin2005} that
cannot be reproduced in Wildcard. We consider this an acceptable
tradeoff in exchange for achieving a gentler slope of customization, and
we show in Section~\ref{sec:examples} and Section~\ref{sec:evaluation}
that our model can still implement many useful customizations in
practice. Also, there is sometimes a way to reframe an imperative script
in terms of our direct manipulation model; for example, a script that
iterates through rows in a page adding some additional information could
be reproduced using a formula in Wildcard. (\emph{todo: mention the fact
that the data abstraction step limits possibilities})

@@

We have used Wildcard to build customizations for many real websites. So
far, most usage has come from members of the project team, although we
have had a few external users try the system. Here we offer our
reflections on using the system, focused on two key questions:

\begin{itemize}
\tightlist
\item
  What is the range of useful customizations possible in this paradigm?
\item
  How feasible is it to build DOM scraping adapters for real websites,
  in practice?
\end{itemize}

\emph{todo}

In order for our idea of third-party customization through Wildcard to
succeed, it is important that creating usable adapters for existing
websites takes minimal effort.

To test the viability of writing DOM scraping adapters, we built
adapters for 10 websites, including transactional sites like Amazon and
Uber Eats, media consumption sites like Hacker News and Youtube, and a
medical results entry platform managed by a member of our research
group. The adapters ranged from 36 to 117 lines of code, averaging 68
lines. In addition, an external developer unaffiliated with the project
contributed one adapter, designed to sort the Github page listing a
user's repositories, and described the experience as ``very
straightforward.''

Some of the challenges of writing a DOM scraping adapter are the same
ones involved with writing normal web scraping code, but the more
interactive nature of Wildcard introduces additional challenges. One
challenge is triggering updates to the spreadsheet data in response to
UI changes that happen after initial page load. Site adapters are
responsible for recognizing these changes by observing the DOM. So far,
we have been able to use event listeners and the MutationObserver API to
successfully observe changes, but it may prove challenging to observe
changes on some sites only through the DOM. Another challenge is
persisting updates to the DOM---some websites use virtual DOM frameworks
that can occasionally overwrite changes made by Wildcard. Despite these
challenges, we've managed to create an implementation that works around
these issues in practice, and results in a usable customization
experience for all the websites we've tried so far.

@@

This paper extends a workshop paper by Litt and Jackson \citep{litt2020}
which presented an earlier prototype version of the Wildcard system. We
build on their work in this paper by rearchitecting the internal
implementation of the system around the table adapter abstraction,
evaluating it on many more websites, and by more fully characterizing
the behavior and design of the system.

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Customizing Software by Direct Manipulation of Tabular Data}

@@

\begin{table*}[]
\begin{tabular}{llrl}
\hline
\textbf{Website} & \textbf{Description} & \textbf{LOC} & \textbf{Example customizations}                                                              \\ \hline
Airbnb           & Travel               & 73                                       & Add Walkability Scores to listings. Sort listings by price.                           \\
Amazon           & Online shopping      & 99                                       & Sort used book sellers by total price, including delivery fees                                \\
Blogger          & Blogging             & 36                                       & Use alternate text editor to edit blog posts                                                 \\
Expedia          & Travel               & 41                                       & Use alternate datepicker to enter travel dates                                               \\
Flux             & Data portal          & 67                                       & Use Wildcard as a faster table editor for editing lab results                                \\
Github           & Code repository      & 62                                       & Sort a user's code repositories by stars to find popular work                                \\
Hacker News      & News                 & 69                                       & Add read times to links. Filter out links that have been read. \\
Instacart        & Grocery delivery     & 48                                       & Sort groceries by price and category. Take notes on items.                                   \\
Uber Eats        & Food delivery        & 117                                      & Sort/filter restaurants by estimated delivery ETA and price.                                 \\
Weather.com  & Weather              & 51                                       & Sort/filter hourly weather to find nice times of day.                                        \\
Youtube          & Videos               & 80                                       & Sort/filter videos by length, to find short videos to watch.                                 \\ \hline
\end{tabular}
\caption{A list of table-driven customizations that we have implemented using Wildcard.}
\label{tab:websites}
\end{table*}

To evaluate table-driven customization in practice, we built the
Wildcard browser extension, which implements table-driven customization
in the context of existing websites. It is built in Typescript, and
works across three major browsers: Chrome, Firefox, and Edge.

We have used Wildcard to customize 11 websites, including transactional
sites like Amazon and Uber Eats, and media consumption sites like Hacker
News and Youtube. Table 1 summarizes all these customizations, including
the number of lines of code in the adapter configuration for each site.

So far, most usage of Wildcard has come from members of the project
team. Here we offer our reflections on using the system, focused on two
key questions:

\begin{itemize}
\tightlist
\item
  How broad is the range of possible customizations in this paradigm?
\item
  How feasible is it to build DOM scraping adapters for real websites,
  in practice?
\end{itemize}

We have found that table-driven customization can serve a broad range of
useful purposes. Here we expand on some examples that illuminate various
aspects of using the system in practice.

\hypertarget{sortingfiltering}{%
\subsubsection{Sorting/filtering}\label{sortingfiltering}}

It might seem that most websites already have adequate sorting and
filtering functionality, but we have found it surprisingly helpful to
add new sorting/filtering functionality to websites using Wildcard.

Sometimes, websites have opaque algorithms which they use to rank
results, which presumably maximize profit but don't offer much control
to the user. For example, Airbnb doesn't allow users to sort listings by
price, and Youtube doesn't allow users to sort the homepage by watch
time. Wildcard enables users to take back some control over this
process.

In other cases we have found that a lack of sorting options seems less
intentional, and simply a missing feature. For example, the Instacart
grocery delivery service doesn't allow users to sort their grocery cart
by price or category. Our guess is that they simply haven't gotten
around to implementing the feature yet.

In the current implementation of Wildcard, users can only sort and
filter entries which are shown on the current page, which means that
users are not entirely liberated from the suggestions of the opaque
algorithm. We could work around this restriction in the future by
fetching content across multiple pages, but we've also found that
sorting/filtering a single page of a paginated list is often an
acceptable outcome, and even a preferable one. For example, it's more
useful to sort 30 recommended Youtube videos than to try to sort all
videos on Youtube; similarly it's convenient to view the Hacker News
front page sorted by read time, rather than all Hacker News articles.

\hypertarget{annotating}{%
\subsubsection{Annotating}\label{annotating}}

Many other web annotation systems focus on annotating text or arbitrary
webpage content, but Wildcard only allows for annotating structured
objects extracted by an adapter, which results in quite a different set
of use cases.

Annotating with Wildcard has proven most useful when taking notes on a
list of possible options: e.g., evaluating possible Airbnb locations to
rent. We have also used it with Instacart's online grocery cart, for
jotting down notes like ``should we get more milk?''

\hypertarget{formulas}{%
\subsubsection{Formulas}\label{formulas}}

Formulas are the most powerful part of the Wildcard system. So far, we
have only built a small number of functions into the language, but by
adding more functions, the fundamental computing model could support a
very broad range of useful computations, as shown by the success of
spreadsheets.

We have found that formulas are useful for fetching data from Web APIs.
We've used them to augment Airbnb listings with walkability scores, and
augment Hacker News articles with estimated read times. One challenge of
the current language design is that supporting each web API requires
adding a new function to the language using code, because web APIs
typically return complex JSON data structures which don't make sense to
display in a single table cell. Adding functions for handling JSON data
to the formula language might help resolve this.

We have also found instances where simple data manipulation is useful:
adding together different subtotals into a total price, or manipulating
the results of a formula expression with arithmetic or concatenating
string labels, as shown in the example in Section~\ref{sec:examples}.
(\emph{todo: this makes sense iff we switch the examples sectino to HN})

\hypertarget{cell-editors}{%
\subsubsection{Cell editors}\label{cell-editors}}

(\emph{note: written with the assumption that Expedia gets cut from the
Examples section})

``Cell editors'' are UI widgets that expose a custom editing UI for a
single cell of the table view. A programmer building a cell editor only
needs to integrate it with the table viewer; actually propagating values
into the website UI is handled by the site-specific DOM adapter. As a
result, creating a new cell editor can be as easy as just creating glue
code between the table editor and an existing UI widget library.

One benefit of cell editors is that a user can use their personal
information in a web UI without uploading it to the website. To explore
this idea, we created a cell editor based on the FullCalendar Javascript
plugin, which can load data from a Google Calendar. This makes it
convenient to enter dates into a website based on the user's personal
calendar information.

Another benefit of cell editors is that a user can choose their
preferred widget for editing some type of information. We built a cell
editor based on the CKEditor rich text editor. As an example of using
this editor, we integrated it with the Blogger website and used it to
edit blog posts, in place of the Blogger website's built-in editor.

\hypertarget{limitations}{%
\subsubsection{Limitations}\label{limitations}}

There are many useful customizations which are not possible in the
table-driven customization paradigm. Wildcard can only make
customizations that use the available data exposed in the table. UI
modifications are limited in scope; for example, deleting arbitrary
buttons isn't possible. At one point, we wanted to build an automation
to repeatedly load a grocery delivery website to check for open delivery
slots, but it wasn't clear how to achieve this. Some of these
limitations are specific to the current implementation of the Wildcard
extension, but some are more general to the entire idea of table-driven
customization.

We consider this an acceptable outcome---our goal is simply to make many
useful customizations available to end users with a low threshold, not
to span all possible customizations. Section~\ref{sec:dm} discusses this
point further.

One benefit of Wildcard is \emph{predictability}: once we built a site
adapter for a website, it was generally obvious to us what types of
customizations were and weren't possible, and the UI guided us towards
building customizations that matched the system's model. For example,
it's clear that it's not possible to customize a website using data not
available in the table.

Our second evaluation area is less about the conceptual approach of
table-driven customization, and more about our specific implementation
of customizing existing web applications. In order for third-party
customization through Wildcard to succeed, it is important that creating
usable adapters for existing websites takes minimal effort.

Most of our DOM scraping adapters were created by members of our team.
However, an external developer unaffiliated with the project contributed
one adapter, designed to sort the Github page listing a user's
repositories. They described the experience as ``very straightforward.''

The adapters for our test sites ranged from 36 to 117 lines of code,
averaging 68 lines; Table 1 shows the full lines of code for each
adapter.

Some of the challenges of writing a DOM scraping adapter are the same
ones as with writing normal web scraping code, but the more interactive
nature of Wildcard introduces additional challenges. rve changes, but it
may prove challenging to observe changes on some sites only through the
DOM. One challenge is triggering updates to the spreadsheet data in
response to UI changes that happen after initial page load. Site
adapters are responsible for recognizing these changes by observing the
DOM. So far, we have been able to use event listeners and the
MutationObserver API to successfully obse

Another challenge is persisting updates to the DOM---some websites use
virtual DOM frameworks that can occasionally overwrite changes made by
Wildcard.

So far, we've managed to work around these issues for all the websites
we've tried, but we don't claim that any website can be customized
through DOM scraping. As web frontend code gets increasingly complex
(and starts to move beyond the DOM to other technologies like Shadow DOM
or even WebGL for rendering), it may become increasingly difficult to
customize websites from the outside without first-party support.

@@

\hypertarget{sec:dm}{%
\subsection{Customization by direct manipulation}\label{sec:dm}}

@@

It is important to note that there are many specific customizations that
can be achieved in systems like Chickenfoot \citep{bolin2005} that
cannot be reproduced in Wildcard. We consider this an acceptable
tradeoff in exchange for achieving a gentler slope of customization, and
we show in Section~\ref{sec:examples} and Section~\ref{sec:evaluation}
that our model can still implement many useful customizations in
practice. Also, there is sometimes a way to reframe an imperative script
in terms of our direct manipulation model; for example, a script that
iterates through rows in a page adding some additional information could
be reproduced using a formula in Wildcard.

@@

@@

This paper extends work reported in a workshop paper by Litt and Jackson
\citep{litt2020} which presented an earlier prototype version of the
Wildcard system. We build on their work in this paper by rearchitecting
the internal implementation of the system around the table adapter
abstraction, evaluating it on many more websites, and by more fully
characterizing the behavior and design of the system.
